---
title: "Mock Microbial Comparison Results"
author: "Caleb Easterly"
date: "October 30, 2018"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
```

# Get True Results
The true protein amounts were expanded with metaQuantome. 
```{r}
true_expanded <- read.delim('true_composition/mqome_outputs/true_out.tab')
```

# Estimate Intensity With Summarization

```{r}
library(plyr)
library(dplyr)
flash_results <- read.delim('estimated_composition/mqome_inputs/flash.tab',
							stringsAsFactors = FALSE)
unipept_results <- read.delim('estimated_composition/mqome_inputs/tax.tab',
							  stringsAsFactors = FALSE)
taxa_with_int <- left_join(unipept_results, flash_results, by = "peptide") %>%
	select(lca, starts_with("Run"))
taxa_with_int$mean_int <- rowMeans(taxa_with_int[, -1], na.rm = TRUE)

summarization_estimates <- taxa_with_int %>%
	group_by(lca) %>%
	summarise(int = sum(mean_int, na.rm=TRUE))
```

# Estimate Intensity with metaQuantome

```{r}
# read metaquantome results
metaquantome_results <- read.delim('estimated_composition/mqome_outputs/tax_out.tab')

library(reshape2)

true <- read.delim('true_composition/mqome_outputs/true_out.tab')
# original_abund <- read.delim('true_composition/true.tab')
```


```{r}
joined_mqome <- left_join(true, metaquantome_results, by = "taxon_name",
						  suffix = c(".true", ".est_int")) %>%
	select(taxon_name, true_abund = g1_mean.true, int_abund = g1_mean.est_int)

joined_orig <- left_join(true, summarization_estimates, by = c("taxon_name" = "lca")) %>%
	mutate(logint = log2(int)) %>%
	select(taxon_name, true_abund = g1_mean, lca_sum_abund = logint)

all_together <- join_all(list(joined_mqome, joined_orig), by = c("taxon_name", "true_abund"))
```

### Scale the dataframe

The true abundances and the estimated abundances are all on different scales, so I use `scale` to ensure they all have a mean of 0 and an standard deviation of 1.

```{r}
scaled <- data.frame(taxon_name = all_together$taxon_name,
					 scale(all_together[, 2:4]))
# colMeans(scaled[, -1], na.rm = TRUE)
# apply(scaled[, -1], 2, function(x) sd(x, na.rm=T))
```

### Mean of Squared Errors

Now that I've scaled the result, I calculate the average distance from the true abundances. I use average because there is a different number of non-missing intensities for spectral counts and precursor intensities.

Intensity:
```{r}
int_err <- with(scaled, mean((true_abund - int_abund)^2, na.rm = TRUE))
```

Original Intensity:
```{r}
lca_err <- with(scaled, mean((true_abund - lca_sum_abund)^2, na.rm = TRUE))
```

```{r}
errs <- data.frame("analysis" = c("metaquantome", "lca (intensity)", "true"),
				   "ntaxa" = c(sum(!is.na(scaled[, "int_abund"])),
				   			sum(!is.na(scaled[, "lca_sum_abund"])),
				   			nrow(scaled)),
				   "mse" = c(int_err, lca_err, '-'))
library(knitr)
library(kableExtra)
errs %>%
	kable() %>%
	kable_styling()
```


<!-- ### Plotting -->

<!-- ```{r} -->
<!-- library(ggplot2) -->
<!-- ms1 <- "Precursor Intensity" -->
<!-- spec <- "Spectral Counting" -->
<!-- lca_sum <- "Sum on LCA" -->
<!-- ggplot(scaled) + -->
<!-- 	geom_point(aes(x = true_abund, y = int_abund, col = ms1)) + -->
<!-- 	geom_point(aes(x = true_abund, y = mq_sc_abund, col=spec)) + -->
<!-- 	geom_abline(intercept = 0, slope = 1) + -->
<!-- 	geom_smooth(aes(x = true_abund, y = int_abund, col = ms1), method = "lm", se=FALSE) + -->
<!-- 	geom_smooth(aes(x = true_abund, y = mq_sc_abund, col = spec), method = "lm", se=FALSE) + -->
<!-- 	geom_smooth(aes(x = true_abund, y = lca_sum_abund, col = lca_sum), method = "lm", se=FALSE) + -->
<!-- 	theme_bw() + -->
<!-- 	labs(x = "Scaled True Abundance", y = "Scaled Estimated Abundance") -->
<!-- ``` -->